---
output:
  pdf_document: default
  html_document: default
---
Some packages you may need to install first:

- optmatch: `install.packages("optmatch")`
- sandwich: `install.packages("sandwich")`
- MatchIt: `install.packages("MatchIt")`
- marginaleffects: `install.packages("marginaleffects")`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("MatchIt")
library("marginaleffects")
```

# Matching with job training data from ``Evaluating the econometric evaluations of training programs with experimental data'' (LaLonde 1986)
We'll use a portion of the data from a job training program. In particular, the treatment is whether or not someone participated in a job training program. The outcome of interest is their salary in 1978 (re78).
```{r}
# Load the data
data("lalonde")

# See what's in the data
?lalonde # this opens up the "Help" tab with the documentation! 
head(lalonde)

```

Now let's try to run the matching procedure using the \texttt{matchit} function. 
```{r}
# method: nearest (i.e. greedy 1:1)
# distance: glm specifies propensity score matching using the propensity score estimated
  # using logistic regression 
# data: lalonde (the data set we're working with)
# replace: True (T) or False (F) - whether to sample with or without replacement. 
    # Note, if allowing for replacement, greedy and optimal are the same
    # So for the function, you only need to specify if using method = "nearest"
# ratio: how many control matches for each treated unit
m.out0 <- matchit(treat ~ re74 + re75 + age + race + educ + married + nodegree,
                  method = "nearest", distance = "glm",
                  data = lalonde, replace = F, 
                  ratio = 1)

```

# Assessing the matching
We can check how well the balancing has been done
```{r}
?summary.matchit # Look in the Help tab (on the bottom right) for documentation on summary.matchit


# interactions: check interaction terms too? (T or F)
# un: show statistics for unmatched data as well (T or F)
summary(m.out0, interactions = F, un = F)
# Std. Mean Diff (SMD): difference of means, standardized by sd of treatment group
# Var. Ratio: ratio of variances in treatment and control group. Compares spread of data
# Rubin (2001) presents rule of thumb that SMD should be less than .25 and variance ratio should be between .5 and 2
# Max eCDF: Kolmogorov-Smirnov statistic. Max difference across entire CDF
```

We can also visually asses the matches
```{r}
## Produces QQ plots of all variables in which.xs
plot(m.out0, type = "qq", interactive = F)
## Plots the density of all variables in which.xs
plot(m.out0, type = "density", interactive = F)
## Plots the empirical CDF of all variables in which.xs
plot(m.out0, type = "ecdf", which.xs = ~age + re74, interactive = F)
## Plots propensity scores of matched vs unmatched
plot(m.out0, type = "jitter")
```
Let's try using calipers
```{r}

m.out0 <- matchit(treat ~ re74 + re75 + age + race + educ + married + nodegree,
                  method = "nearest", distance = "glm",
                  data = lalonde, replace = F, caliper = .05, std.caliper = F,
                  ratio = 1)
summary(m.out0, interactions = F, un = F)

```

As an alternative, let's try with coarsened exact matching
```{r}
# method: nearest (i.e. greedy 1:1)
# distance: glm specifies propensity score matching using the propensity score estimated
  # using logistic regression 
# data: lalonde (the data set we're working with)
# replace: True (T) or False (F) - whether to sample with or without replacement. 
    # Note, if allowing for replacement, greedy and optimal are the same
    # So for the function, you only need to specify if using method = "nearest"
# ratio: how many control matches for each treated unit
m.out1 <- matchit(treat ~ re74 + re75 + age + race + educ + married + nodegree,
                  method = "cem",
                  data = lalonde, replace = F, 
                  ratio = 1)

# We only match 65 of the original 185 treated units
summary(m.out1, interactions = F, un = F)

m.out2 <- matchit(treat ~ re74 + re75 + age + race + educ + married + nodegree,
                  method = "cem", 
                  grouping = list(race = list(c("black"), c("hispan", "white"))),
                  data = lalonde, replace = F, 
                  ratio = 1)

# We only match 65 of the original 185 treated units
summary(m.out2, interactions = F, un = F)

# We could group hispanic/white together for the purposes of matching
# and not enforce matching on married, educ, nodegree
m.out2 <- matchit(treat ~ re74 + re75 + age + race,
                  method = "cem", 
                  grouping = list(race = list(c("black"), c("hispan", "white"))),
                  data = lalonde, replace = F, 
                  ratio = 1)
summary(m.out2, interactions = F, un = F, addlvariables = ~ married + nodegree + educ)

match.data(m.out2) %>%
  arrange(subclass)


```


# Estimating the effect
Given the matching (and assuming it is good enough), we can estimate the ATT. But we can do even better by combining regression + matching.
```{r}

# propensity score matched
m.out0.data <- match.data(m.out0)
# CEM matched
m.out2.data <- match.data(m.out2)

# Fitting a linear model on the treatments will work
# even if all weights are not 1. We just need to feed them in
fit1 <- lm(re78 ~ treat, data = m.out2.data, weights = weights)

# vcov: tells R to use robust standard errors
avg_comparisons(fit1, variables = "treat",
                vcov = "HC3",
                newdata = subset(m.out2.data, treat == 1),
                wts = "weights")


# Fitting a linear model on the treatments will work
# even if all weights are not 1. We just need to feed them in
fit2 <- lm(re78 ~ treat + re74 + re75 + age + race + married + nodegree + educ,
           data =m.out2.data, weights = weights)

# vcov: tells R to use robust standard errors
avg_comparisons(fit2, variables = "treat",
                vcov = "HC3",
                newdata = subset(m.out2.data, treat == 1),
                wts = "weights")


fit3 <- lm(re78 ~ treat + re74 + re75 + age + race + married + nodegree + educ,
           data =m.out0.data, weights = weights)

# vcov: tells R to use robust standard errors
avg_comparisons(fit3, variables = "treat",
                vcov = "HC3",
                newdata = subset(m.out0.data, treat == 1),
                wts = "weights")

```


