---
output: pdf_document
---
# Discussion 2. Analyzing an Experiment in R


### Necessary packages

Note: If this errors you probably either don't have `dplyr`, `tidyr` or `readr`
installed. Use `install.packages("package_name")` to install `package_name`

```{r}
knitr::opts_chunk$set(eval = FALSE)

library(dplyr)
library(tidyr)
library(readr)
```

### Import data
```{r}
genAI_dta <- read_csv("https://raw.githubusercontent.com/causal3900/causal3900.github.io/refs/heads/main/assets/discussions/discussion02_files/GenAI_Learning.csv")
```


Run the following code to get a quick peek at the dataset using the function `glimpse`. This returns info such as the number of rows/columns, the column names, and the type of data in each column. Notice that we have information about honors class `Honors`. Also notice that the treatment variables `GPTBase` and `GPTTutor`.

```{r}
glimpse(genAI_dta)
```

### Clean data

First, let's exclude honors-designated classrooms from the sample. The `Honors` variable says if the student is in an Honors class. For this, we use the `filter` function.

- The general syntax is `filter(condition)`
- A condition would be `Honors == 0`

We have the code started for you below. Fill in the appropriate condition to filter by:

```{r}
genAI_dta <- genAI_dta %>%
  filter(...)
```

Now, create the `treatment` variable from the variables `GPTBase` and `GPTTutor` with the labels: 
- "Control": If `GPTBase == 0` and `GPTTutor == 0`
- "GPTBase": If `GPTBase == 1` and `GPTTutor == 0`
- "GPTTutor": If `GPTBase == 0` and `GPTTutor == 1`

For this, you will want to use the function `case_when`. The general syntax is `case_when(condition ~ output-value)`. For example, a condition would be `(GPTBase == 0) & (GPTTutor == 0)` and an output value would be `"Control"`. This would search for every value in the `GPTBase` and `GPTTutor` columns that equal `0` and replace that with the string `"Control"`.


We have started the code for you below. Decide what argument(s) to pass inside the parentheses of `case_when()`.

```{r}
genAI_dta <- genAI_dta %>%
  mutate(treatment = case_when(...)) 
```


### Balance table

Next, we're going to confirm that our control and treatment groups look pretty
much the same across a set of covariates, i.e. that the two groups are *balanced on covariates*. Specifically this means we're going to calculate the mean value of a set of covariates across each of the treatment/control arms, and we expect them to be pretty much equal if our randomization worked. This is related to the idea of exchangeability. 

In this exercise, we are going to reproduce a table similar to Table A.1 from the appendix. We want a table that shows the mean value of the covariates given below. Notice that some of these covariates are binary (i.e. 0/1), what would the mean value of those variables stand for? You should create a table with 3 columns, one for each treatment arm, and 13 rows, one for each covariate of interest. 

We have started some code for you below. What you need to do is:

- Pass an argument to `group_by()` so that we calculate separate means for each treatment arm.
- Pass an argument to `summarise()` that computes the mean of each covariate in `covariates` for each separate treatment arm.

```{r}
covariates <- c("education_parent_college", "n_household_members",
                "n_household_children", "class_enjoyment",
                "class_participation_likelihood", "math_hw_completion", "hw_help",
                "private_tutorship","visit_training_center","female",
                "n_weekday_study_hours","n_weekend_study_hours","gpa_prev")

genAI_balance <- genAI_dta |>
  distinct(`Student ID`, .keep_all = TRUE) |>
  group_by(...) |>
  summarise(across(all_of(...),\(x) mean(x, na.rm = TRUE)))


genAI_balance |>
  pivot_longer(-treatment, names_to = "Description", values_to = "value") |>
  pivot_wider(names_from = treatment, values_from = "value") |>
  select(Description, GPTTutor, Control, GPTBase) |>
  print()

```

What you want to notice is that the values in each row are similar across the columns.

### Results

Finally, let's replicate the main results (Table 1). For each treatment arm, calculate the average grade of students in the:
    - Practice problems (`Part2Tot`)
    - Exam problems (`Part3Tot`)
We started the code for you below. You need to 
- Fill in the `group_by()` so that we are working with each treatment arm seperately
- Fill in two arguments to `mean()` to do the following:
  - Ensure the column titled `Practice_perf` is the average grade in step 2 of each group (`Part2Tot`)
  - Ensure the column titled `Exam_perf` is the the average grade in step 3 of each group (`Part3Tot`)
Note that your numbers will not match up exactly with Table 1


```{r}
genAI_means <- 
  genAI_dta %>%
  group_by(...) %>%
  summarise(Practice_perf = mean(...),
            Exam_perf = mean(...))

print(genAI_means)
```

Next, we will estimate the average treatment effect (ATE). There is nothing to fill in, simply make sure the numbers are roughly as in Table 1 in the paper.

```{r}
ATE_practice <-
  genAI_means$Practice_perf[genAI_means$treatment!="Control"] -
  genAI_means$Practice_perf[genAI_means$treatment=="Control"]

ATE_exam <-
  genAI_means$Exam_perf[genAI_means$treatment!="Control"] -
  genAI_means$Exam_perf[genAI_means$treatment=="Control"]


print(cbind(ATE_practice,ATE_exam))
```
