
# Instrumental variables

## Experimental settings

> Oct 17. [**Slides.**](assets/slides/7-1_iv_experimental.pdf)

An instrumental variable (IV) identification strategy applies when a treatment effect of $A$ on $Y$ is confounded by unobserved variables ($U$), but an **instrument** $Z$ creates random unconfounded variation in $A$.

<script type="text/tikz">
  \begin{tikzpicture}
    \node (l) at (0,0) {$L$};
    \node (a) at (1,0) {$A$};
    \node (y) at (2,0) {$Y$};
    \draw[->] (l) -- (a);
    \draw[->] (a) -- (y);
    \draw[->] (l) to[bend right] (y);
  \end{tikzpicture}
</script>

A very clean setting for IV is randomized experiments with non-compliance: an experimenter randomizes the assigned treatment ($Z$) but the actual treatment ($A$) may be unequal to $Z$ because some units do not follow their assignment. Our first class will discuss this setting.

## Observational settings

> Oct 19 [**Slides.**](assets/slides/7-2_iv_observational.pdf) After class, read [Hernán and Robins 2020](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/) Chapter 16.

On Thursday, we move on to IV analysis in observational settings. Here we focus on the casual assumptions required for IV. These assumptions often hold by design in experiments with non-compliance. In observational settings, they can be more doubtful.

## Lab: Estimation

In lab, between the two lectures, we will implement instrumental variables estimators in R. We will put instructions here for accessing a dataset, but for now the code below generates simulated data.

```{r, message = F, warning = F}
library(tidyverse)
```

```{r}
iv_data <- data.frame(Z = rbinom(1e3,1,.5)) %>%
  mutate(U = rnorm(n()),
         p_A = plogis(-.5 + Z + U),
         A = rbinom(n(), 1, p_A),
         Y = rnorm(n(), U + A)) %>%
  select(-p_A)
```

### Wald estimator

The Wald estimator proceeds by (1) estimating the effect of the instrument on the outcome, (2) estimating the effect of the instrument on the treatment, and (3) dividing (1) / (2).
$$\begin{aligned}
\hat{E}(Y^{a=1}-Y^{a=0}) &= \frac{\hat{E}(Y^{z=1}-Y^{z=0})}{\hat{E}(A^{z=1}-A^{z=0})}\\
&= \frac{\hat{E}(Y\mid Z = 1) - \hat{E}(Y\mid Z = 0)}{\hat{E}(A\mid Z = 1) - \hat{E}(A\mid Z = 0)}
\end{aligned}$$

Implement the Wald estimator in code.

1. Estimate the mean difference in $Y$ over groups defined by $Z$
2. Estimate the mean difference in $A$ over groups defined by $Z$
3. Divide (1) / (2)

### Two-stage least squares

A second estimator for IV is two-stage least squares.

1. Use OLS to model $A$ as a function of $Z$
2. Store predicted values $\hat{A}$ without modifying $Z$
3. Use OLS to model $Y$ as a function of $\hat{A}$

The coefficient on $\hat{A}$ in (3) estimates the effect of $A$ on $Y$.