---
output: html_document
---
<style>li {line-height: 1.8;}</style>

# Discussion 2. Analyzing an Experiment in R {-}
## STSCI/INFO/ILRST 3900: Causal Inference {-}
#### September 2, 2026 {-}



```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(dplyr)
library(tidyr)
library(readr)
```
 

### Announcements {-}
- Office Hours throughout the week (see Syllabus or website)
  - Filippo: Thursday 4-5pm in 325A CIS Building
  - Shira: Monday 5-6 pm in 325A CIS Building
- No need to submit, encouraged to complete the file and knit


### Generative AI and Math Learning Experiment {-}
- How generative AI affects learning?
- GenAI already shown to improve human productivity
- Is GenAI an effective tutor, or does it become a cognitive crutch?
- __Research Question:__ To what extent does access to GenAI (with and without guardrails) affect learning in an educational setting?
- Article: ["Generative AI without guardrails can harm learning: Evidence from high school mathematics"](https://www.pnas.org/doi/pdf/10.1073/pnas.2422633122)
- We’ll be analyzing their experiment today!

### Experimental Design {-}
- Nearly 1,000 high school students in Turkey were randomly assigned 1 of 3
- High school math students were randomly assigned to one of three treatment arms:
  1. **Control:** Standard practice material with no AI assistance access
  2. **GPT Base:** Access to standard GPT-4 
  3. **GPT Tutor:** Access to a customized GPT-4 (GenAI with guardrails)
-  Following this practice session, all students took an exam without AI access to evaluate learning
 

### Goal for Today {-}
__Replicate Something Similar to Table A.1 (in appendix) and Table 1__
- Table A.1 demonstrates balance of covariates across treatment arms
- This ensures treatment assignment is randomized so exchangeability holds
- Table 1 shows the main results of student performance in the practice and exam problems
- Estimated the counterfactuals to identify the average causal effect (ACE) 
$$ACE = E[Y^\text{a=1}]- E[Y^\text{a=0}]$$

### Resources for Markdown {-}
- Hadley Wickham's R for Data Science ["Chapter 27"](https://r4ds.had.co.nz/r-markdown.html)
- ["RMarkdown cheat sheet"](https://rmarkdown.rstudio.com/lesson-15.HTML) from RStudio
- ["Data Wrangling and Analyses with Tidyverse"](https://bookdown.org/content/24942ad6-9ed7-44e9-b214-1ea8ba9f0224/data-wrangling-and-analyses-with-tidyverse.html) by Bookdown
- ["RMarkdown for Scientists"](https://rmd4sci.njtierney.com/) by Nicholas Tierney
- If you can’t figure out how to do something, try Googling it first!
- Also feel free to ask a classmate or ask me :)


### Step 1: Download the [**.Rmd file here**](assets/discussions/discussion2.Rmd) {-}
- Start by running the code in Section “Necessary packages”
- If you get an error, you may need to install the package

### Step 2: Import and Clean the Data {-}

```{r}
genAI_dta <- read_csv("https://raw.githubusercontent.com/causal3900/causal3900.github.io/refs/heads/main/assets/discussions/discussion02_files/GenAI_Learning.csv")

```

- Quick peek at the dataset using the [function `glimpse`](https://dplyr.tidyverse.org/reference/glimpse.html)
- Notice the variable `Honors`
- Notice the treatment variables `GPTBase` and `GPTTutor`

```{r}
glimpse(genAI_dta)
```


- _...students are randomly assigned to classrooms (with the exception of honors-designated classrooms)..._
- Exclude honors-designated classrooms from the sample
- For this, you will want to use the function `filter`
- The general syntax is `filter(condition)`
- A condition would be `Honors == 0`


```{r}
genAI_dta <- genAI_dta %>%
  filter()
```


- Create a single treatment variable with word labels: 
  - "Control": If `GPTBase == 0` and `GPTTutor == 0`
  - "GPTBase": If `GPTBase == 1` and `GPTTutor == 0`
  - "GPTTutor": If `GPTBase == 0` and `GPTTutor == 1`

- For this, you will want to use the function `case_when` which is described [here](https://www.sharpsightlabs.com/blog/case-when-r/)
- The general syntax is `case_when(condition ~ output-value)`
- A condition would be `(GPTBase == 0) & (GPTTutor == 0)` and an output value would be `"Control"`

```{r eval=FALSE}
genAI_dta <- genAI_dta %>%
  mutate(treatment = case_when()) 
```


### Step 3: Table A.1 {-}
- Is the data balanced on covariates?
- We want to check that the treatment groups are balanced on covariates
- For each treatment arm/group, calculate the mean for each of the designated covariates in table A.1 (list given below)
- Use [`group_by()`](https://dplyr.tidyverse.org/reference/group_by.html) to calculate separate means for each treatment arm
- Use [`summarise()`](https://dplyr.tidyverse.org/reference/summarise.html)  to compute the mean of each covariate in `covariates`


```{r eval=FALSE}
covariates <- c("education_parent_college", "n_household_members",
                "n_household_children", "class_enjoyment",
                "class_participation_likelihood", "math_hw_completion", "hw_help",
                "private_tutorship","visit_training_center","female",
                "n_weekday_study_hours","n_weekend_study_hours","gpa_prev")

genAI_balance <- genAI_dta |>
  distinct(`Student ID`, .keep_all = TRUE) |>
  group_by() |>
  summarise(across(all_of(),\(x) mean(x, na.rm = TRUE)))


genAI_balance |>
  pivot_longer(-treatment, names_to = "Description", values_to = "value") |>
  pivot_wider(names_from = treatment, values_from = "value") |>
  select(Description, GPTTutor, Control, GPTBase) |>
  print()

```


### Step 4: Table 1 {-}
- What are the main results of the experiment?
- Estimate the regression model for the practice performance
$$\small{\text{Outcome}_{ics}=\beta_1\text{GPTBase}_c+\beta_2\text{GPTTutor}_c+\beta_3\text{prevGPA}+\theta_s+\delta_g+\alpha_y+\lambda_t+\varepsilon_{ics}}$$
 where $\text{Outcome}_{ics}$ is the normalized grade of student $i$ in classroom $c$ and session $s$ for the practice problems: `Part2Tot`
 - For this, you will want to use the function `lm` which is described [here](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/lm)
- The general syntax is `lm(outcome ~ independent_variables)`

```{r eval=FALSE}
reg_practice <- lm( ~ GPTBase + GPTTutor + gpa_prev +
                     teacher + Session + Grader + Year, data = genAI_dta)
reg_practice %>%
  summary() %>%
  coefficients() %>%
  head(4)
  

reg_exam <- lm( ~ GPTBase + GPTTutor + gpa_prev +
                 teacher + Session + Grader + Year, data = genAI_dta)
reg_exam %>%
  summary() %>%
  coefficients() %>%
  head(4)

```

- Estimate potential outcomes under each treatment arm
- First, create three data frames where all students receive the same treatment
- For this, you can use `mutate()` function to specify the treatment
- Next, use `predict()` to estimate the potential outcomes- i.e. fitted values for the new data frames
- Finally, use `mean()` to estimate the ACE
$$\widehat{ACE} = \frac{1}{n}\sum_{i=1}^n \hat Y^\text{a=1} - \frac{1}{n}\sum_{i=1}^n \hat Y^\text{a=0}$$

```{r}
genAI_dta_control <- genAI_dta %>%
  mutate(GPTBase = ,
         GPTTutor = )

genAI_dta_GPTBase <- genAI_dta %>%
  mutate(GPTBase = ,
         GPTTutor = )

genAI_dta_GPTTutor <- genAI_dta %>%
  mutate(GPTBase = ,
         GPTTutor = )

y_control <- predict(reg_practice,newdata = , na.action = na.omit)
y_Base <- predict(reg_practice,newdata = , na.action = na.omit)
y_Tutor <- predict(reg_practice,newdata = , na.action = na.omit)


(ACE_Base  <- mean() - mean())
(ACE_Tutor  <- mean() - mean())
```



