<style>li {line-height: 1.8;}</style>

# Discussion 4. Interference and Stat Review {-}
## STSCI/INFO/ILRST 3900: Causal Inference {-}
#### September 17, 2025 {-}


## Interference {-}
You can download the [**slides**](assets/discussions/discussion4-interference.pdf) for this week's discussion.

```{r message=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
```

## Stat review reminder {-}
- $\pi^a= P(Y^a=1)$) under treatment condition $a$
- Let $\hat\pi^a$ be the estimate of that unknown probability
- $\hat\pi^a = \frac{1}{n_a}\sum_{i:A_i=a} Y_i^a$, proportion of people in that treatment condition to have outcome 1
- To make a confidence interval on $\hat\pi^a$ we need the standard error of $\hat\pi^a$

### Standard error {-}
- Let $Y^a$ be a Bernoulli random variable
- The variance of $Y^a$ is $V(Y^a) = \pi^a (1-\pi^a)$
- We have estimated by an average: $\hat\pi^a$
- If we did this many times in many hypothetical samples, we would get different estimate
- The estimates sampling variance is $V(\hat\pi^a) = \frac{\pi^a(1-\pi^a)}{n_a}$
- The standard error is the square root of the sampling variance: 
$$SE(\hat\pi^a) = \sqrt\frac{\pi^a(1-\pi^a)}{n_a}$$
- We can estimate this SE by plugging in our estimate $\hat\pi^a$
- In R: we translated the standard error formula into code


> ```{r}
se_binary <- function(p, n) {
  se <- sqrt( p * (1 - p) / n )
  return(se)
}
> ```

### Sampling distribution {-}

- $\hat\pi^a$ is a sample mean
- By the Central Limit Theorem: as $n \to \infty$, across hypothetical repeated samples the distribution of $\hat\pi^a$ estimates becomes Normal
- Across repeated samples, the middle 95% of estimates will fall within a known range:
$$\pi^a \pm \Phi^{-1}(.975) \times SE(\hat\pi^a)$$
- where $\Phi^{-1}()$ is the quantile of the standard Normal distribution
- $\Phi^{-1}(.975) \approx 1.96$, so the number 1.96 might be familiar to you 

```{r, echo = F, fig.height = 2}
normal_curve <- data.frame(x = qnorm(seq(.01,.99,.0025))) %>%
  mutate(f = dnorm(x))

normal_curve %>%
  ggplot(aes(x = x, y = f)) +
  geom_ribbon(data = normal_curve %>%
                filter(x >= qnorm(.025) & x <= qnorm(.975)),
              aes(ymin = 0, ymax = f),
              color = "gray", alpha = .2) +
  geom_line() +
  geom_hline(yintercept = 0) +
  annotate(geom = "segment", 
           x = qnorm(c(.025,.5,.975)), 
           xend = qnorm(c(.025,.5,.975)),
           y = 0, yend = dnorm(qnorm(c(.025,.5,.975))),
           linetype = "dashed") +
  annotate(geom = "text", hjust = 0,
           x = 0.05, y = .7 * dnorm(0),
           label = "True\nParameter\nValue") +
  annotate(geom = "segment",
           x = qnorm(.1), xend = qnorm(.9),
           y = .2 * dnorm(0), yend = .2 * dnorm(0),
           arrow = arrow(ends = "both", length = unit(8,"pt"))) +
  annotate(geom = "label",
           x = 0, y = .2*dnorm(0),
           label = "'Middle 95% of Sample-Based Estimates'~hat(pi)^a",
           parse = T,
           fill = "gray", label.size = NA) +
  annotate(geom = "text", label = "Estimate is normally distributed over\nhypothetical repeated samples by the\nCentral Limit Theorem because\nthe estimate is a sample mean",
           size = 3,
           x = qnorm(.99), y = dnorm(qnorm(.5)),
           hjust = 1, vjust = 1) +
  scale_x_continuous(breaks = c(-qnorm(.975),0,qnorm(.975)),
                     labels = c(expression(pi^a - {Phi^{-1}} (.975)*SE(hat(pi)^a), 
                                           pi^a, 
                                           pi^a + {Phi^{-1}} (.975)*SE(hat(pi)^a)))) +
  theme_void() +
  theme(axis.text.x = element_text(color = "black"))
``` 


- Plug in the estimates $\hat\pi^a$ and $\widehat{SE}(\hat\pi^a)$ to get a 95% confidence interval
- The CI is centered on the estimate $\hat\pi^a$
- If we repeatedly made a CI using hypothetical samples from the population, the CI would contain the unknown true parameter $\pi^a$ 95% of the time
- In R: we translated the confidence interval formula into code


> ```{r}
ci_lower <- function(estimate, standard_error) {
  estimate - qnorm(.975) * standard_error
}
ci_upper <- function(estimate, standard_error) {
  estimate + qnorm(.975) * standard_error
}
> ```
