---
output: html_document
---
<style>li {line-height: 1.8;}</style>

# Discussion 2. Analyzing an Experiment in R {-}
## STSCI/INFO/ILRST 3900: Causal Inference {-}
#### September 2, 2026 {-}



```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(dplyr)
library(tidyr)
library(readr)
```
 

### Announcements {-}
- Office Hours throughout the week (see Syllabus or website)
  - Filippo: Thursday 4-5pm in 325A CIS Building
  - Shira: Monday 5-6 pm in 325A CIS Building
- No need to submit, encouraged to complete the file and knit
- In-class assignments will start later in the semester


### Generative AI and Math Learning Experiment {-}
- How generative AI affects learning?
- GenAI already shown to improve human productivity
- Is GenAI an effective tutor, or does it become a cognitive crutch?
- __Research Question:__ To what extent does access to GenAI (with and without guardrails) affect learning in an educational setting?
- Article: ["Generative AI without guardrails can harm learning: Evidence from high school mathematics"](https://www.pnas.org/doi/pdf/10.1073/pnas.2422633122)
- We’ll be analyzing their experiment today!

### Experimental Design {-}
- Nearly 1,000 high school students in Turkey were randomly assigned 1 of 3
- High school math students were randomly assigned to one of three treatment arms:
  1. **Control:** Standard practice material with no AI assistance access
  2. **GPT Base:** Access to standard GPT-4 
  3. **GPT Tutor:** Access to a customized GPT-4 (GenAI with guardrails)
-  Following this practice session, all students took an exam without AI access to evaluate learning
 

### Goal for Today {-}
__Replicate Something Similar to Table A.1 (in appendix) and Table 1__

- Table A.1 demonstrates balance of covariates across treatment arms
- This ensures treatment assignment is randomized so exchangeability holds
- Table 1 shows the main results of student performance in the practice and exam problems
- Identify the average Treatment effect (ATE) utilizing randomization
$$ACE = E[Y^\text{a=1}]- E[Y^\text{a=0}] = E[Y \mid A=1]-E[Y\mid Y=0]$$

### Resources for Markdown {-}
- Hadley Wickham's R for Data Science ["Chapter 27"](https://r4ds.had.co.nz/r-markdown.html)
- ["RMarkdown cheat sheet"](https://rmarkdown.rstudio.com/lesson-15.HTML) from RStudio
- ["Data Wrangling and Analyses with Tidyverse"](https://bookdown.org/content/24942ad6-9ed7-44e9-b214-1ea8ba9f0224/data-wrangling-and-analyses-with-tidyverse.html) by Bookdown
- ["RMarkdown for Scientists"](https://rmd4sci.njtierney.com/) by Nicholas Tierney
- If you can’t figure out how to do something, try Googling it first!
- Also feel free to ask a classmate or ask me :)


### Step 1: Download the [**.Rmd file here**](assets/discussions/discussion2.Rmd) {-}
- Start by running the code in Section “Necessary packages”
- If you get an error, you may need to install the package

### Step 2: Import and Clean the Data {-}

```{r}
genAI_dta <- read_csv("https://raw.githubusercontent.com/causal3900/causal3900.github.io/refs/heads/main/assets/discussions/discussion02_files/GenAI_Learning.csv")

```

- Quick peek at the dataset using the [function `glimpse`](https://dplyr.tidyverse.org/reference/glimpse.html)
- Notice the variable `Honors`
- Notice the treatment variables `GPTBase` and `GPTTutor`

```{r}
glimpse(genAI_dta)
```


- _...students are randomly assigned to classrooms (with the exception of honors-designated classrooms)..._
- Exclude honors-designated classrooms from the sample
- For this, you will want to use the function `filter`
- The general syntax is `filter(condition)`
- A condition would be `Honors == 0`


```{r}
genAI_dta <- genAI_dta %>%
  filter(...)
```


- Create a single treatment variable with word labels: 
  - "Control": If `GPTBase == 0` and `GPTTutor == 0`
  - "GPTBase": If `GPTBase == 1` and `GPTTutor == 0`
  - "GPTTutor": If `GPTBase == 0` and `GPTTutor == 1`

- For this, you will want to use the function `case_when` which is described [here](https://www.sharpsightlabs.com/blog/case-when-r/)
- The general syntax is `case_when(condition ~ output-value)`
- A condition would be `(GPTBase == 0) & (GPTTutor == 0)` and an output value would be `"Control"`

```{r eval=FALSE}
genAI_dta <- genAI_dta %>%
  mutate(treatment = case_when(...)) 
```


### Step 3: Table A.1 {-}
- Is the data balanced on covariates?
- We want to check that the treatment groups are balanced on covariates
- For each treatment arm/group, calculate the mean for each of the designated covariates in table A.1 (list given below)
- Use [`group_by()`](https://dplyr.tidyverse.org/reference/group_by.html) to calculate separate means for each treatment arm
- Use [`summarise()`](https://dplyr.tidyverse.org/reference/summarise.html)  to compute the mean of each covariate in `covariates`


```{r eval=FALSE}
covariates <- c("education_parent_college", "n_household_members",
                "n_household_children", "class_enjoyment",
                "class_participation_likelihood", "math_hw_completion", "hw_help",
                "private_tutorship","visit_training_center","female",
                "n_weekday_study_hours","n_weekend_study_hours","gpa_prev")

genAI_balance <- genAI_dta |>
  distinct(`Student ID`, .keep_all = TRUE) |>
  group_by(...) |>
  summarise(across(all_of(...),\(x) mean(x, na.rm = TRUE)))


genAI_balance |>
  pivot_longer(-treatment, names_to = "Description", values_to = "value") |>
  pivot_wider(names_from = treatment, values_from = "value") |>
  select(Description, GPTTutor, Control, GPTBase) |>
  print()

```


### Step 4: Table 1 {-}
- What are the main results of the experiment?
- For each treatment arm, calculate the average grade of students in the:
    - Practice problems (`Part2Tot`)
    - Exam problems (`Part3Tot`)
- Use `group_by()` to calculate separate means for each treatment arm
- Use `summarise()` to do the following:
  - Create a column `Practice_perf`- the average grade in step 2 of each group
  - Create a column `Exam_perf`- the average grade in step 3 of each group

- Note that your numbers will not match up exactly with Table 1

```{r eval=FALSE}
genAI_means <- 
  genAI_dta %>%
  group_by(...) %>%
  summarise(Practice_perf = mean(...),
            Exam_perf = mean(...))

print(genAI_means)
```


- Use the sample means to estimate the average treatment effect

```{r eval=FALSE}
ATE_practice <-
  genAI_means$Practice_perf[genAI_means$treatment!="Control"] -
  genAI_means$Practice_perf[genAI_means$treatment=="Control"]

ATE_exam <-
  genAI_means$Exam_perf[genAI_means$treatment!="Control"] -
  genAI_means$Exam_perf[genAI_means$treatment=="Control"]


print(cbind(ATE_practice,ATE_exam))
```



## Example from Lecture {-}

- Given potential outcomes, you chose a diet
- Are $(Y^{a=0},Y^{a=1}) \perp\!\!\!\perp A$?
- Observational setting: no access to potential outcomes, only observed outcome


| | $Y_i = \text{Disease } (0)$ | $Y_i = \text{No Disease } (1)$ |
| :--- | :---: | :---: |
| **$A_i = \text{Delicious } (0)$** | <span style="color:red">20</span> | <span style="color:red">54</span> |
| **$A_i = \text{Healthy } (1)$** | <span style="color:red">0</span> | <span style="color:red">64</span> |


$$\widehat{ATE} = E(Y \mid A = 1) - E(Y \mid A = 0) = \frac{64}{0+64}- \frac{54}{20+54} \approx 0.27$$

### True ATE{-}
- 50% of students have  $Y_i^{\text{Delicious}} = 1$, $Y_i^{\text{Healthy}} = 1$
- 50% of students have $Y_i^{\text{Delicious}} = 0$, $Y_i^{\text{Healthy}} = 1$
- $Y_i^{\text{Delicious}} \sim Bernoulli(0.5)$
- $Y_i^{\text{Healthy}} \sim Bernoulli(1)$

$$ATE = E(Y^{\text{Healthy}}) - E(Y^{\text{Delicious}}) = 1-0.5 = 0.5$$

### Assume a Randomized Trial{-}

```{r eval= TRUE, echo=TRUE}
n <- 138

Y1 <- rep(1,n)
Y0 <- c(rep(0,n/2),
        rep(1,n/2))

set.seed(1231)
A <- rbinom(n, 1, 0.5)

Y <- ifelse(test = (A==1), yes = Y1, no = Y0)

print(table(A,Y))
```
| | $Y_i = \text{Disease } (0)$ | $Y_i = \text{No Disease } (1)$ |
| :--- | :---: | :---: |
| **$A_i = \text{Delicious } (0)$** | <span style="color:red">37</span> | <span style="color:red">41</span> |
| **$A_i = \text{Healthy } (1)$** | <span style="color:red">0</span> | <span style="color:red">60</span> |


$$\widehat{ATE} = E(Y \mid A = 1) - E(Y \mid A = 0) = \frac{60}{0+60}- \frac{41}{41+37} \approx 0.47$$

