---
output: html_document
---

# Discussion 3. Analyzing an Experiment in R {-}

[**Slides**](assets/discussions/discussion3-slides.pdf)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)
```

## Download .Rmd Document

Download today's [**.Rmd document here**](assets/discussions/discussion3.Rmd).


## Get out and Vote Experiment

In this lab, we explore an experiment that digs into the mechanisms
underlying __why people vote__. This exercise is based on: 

Gerber, Alan S., Donald P. Green, and Christopher W. Larimer. ["Social Pressure and Voter Turnout: Evidence from a Large-scale Field Experiment."](https://www.cambridge.org/core/journals/american-political-science-review/article/social-pressure-and-voter-turnout-evidence-from-a-largescale-field-experiment/11E84AF4C0B7FBD1D20C855972C2C3EB#) American Political Science Review 102.1 (2008): 33-48.

A long-standing theory as to why many people
vote is that it is driven by social norms (e.g. the understanding that voting
is their civic duty). This theory, while being a dominant theoretical 
explanation, had very little empirical backing for a long time. This experiment
examines this very theory by asking the question:
__to what extent do social norms cause voter turnout__?

### Experimental Design

In order to answer this question, approximately 80,000 Michigan households
were randomly assigned to treatment and control groups, where the treatment
group was randomly assigned to one of four possible treatment arms. These
treatment arms varied in the intensity of social pressure that they conveyed,
and were defined as follows:

1. The first treatment arm was mailed a letter that simply reminded them that
voting is a civic duty.
2. The second treatment arm was mailed a letter telling them that researchers
would be studying their voting turnout based on public records.
3. The third treatment arm was mailed a letter stating that their voting turnout 
would be revealed to all other members of their household.
4. The fourth treatment arm was mailed a letter stating that their voting turnout
would be revealed to their household *and* to their neighbors.

## Analyze Experiment
Download this RMarkdown file [here]().

### Necessary packages

__Note:__ If this errors you probably either don't have `dplyr` or `haven`
installed.

```{r}
library(dplyr)
library(haven)
```

### Import data
```{r}
gotv <- read_dta("https://causal3900.github.io/assets/data/social_pressure.dta")
```

> Alternatively (if you really want), you could download the data [here](assets/data/social_pressure.dta) and load it directly from your computer. Make sure to save the data into the same directory that your RMarkdown file is in. 
> Then you can you can import the data as:
>
> `gotv <- read_dta("social_pressure.dta")`

Run the following code to get a quick peek at the dataset using the [function `glimpse`](https://dplyr.tidyverse.org/reference/glimpse.html). This returns info such as the number of rows/columns, the column names, and the type of data in each column. Notice that we have information about year of birth `yob` but not explicitly age. Also notice that the treatments are labeled with the numbers 0 through 4.
```{r}
glimpse(gotv)
```

### Clean data

First, let's construct an age variable describing how old (in number of years)
each person was in the year 2006. The `yob` variable says which year each person
was born in. For this, we use the `mutate` function, which you can read about [here](https://dplyr.tidyverse.org/reference/mutate.html). 

> Given a person's year of birth, how do you calculate their age in the year 2006? Note that you can do arithmetic operations with information from the dataset. For example, if you have two columns `col_1` and `col_2` and you wanted to create a third column called `col_3` that was the sum of these two columns, you could write:
>
> `mutate(col_3 = col_1 + col_2)`

We have the code started for you below. Fill in the appropriate expression after `age = ` to add a column to `gotv` labeled `age` that contains how old each person was in 2006.

```{r}
gotv <- gotv |>
  mutate(age = )
```

Now, convert the `treatment` variable from it's numeric representation to the
corresponding labels which are

- 0: "Control" 
- 1: "Hawthorne" (this is the 'researchers viewing records via public data' treatment arm)
- 2: "Civic Duty" (this is the 'voting is your civic duty' treatment arm)
- 3: "Neighbors" (this is the 'voting turnout revealed to neighbors' treatment arm)
- 4: "Self" (this is the 'voting turnout revealed to household' treatment arm)


> For this, you will want to use the function `case_when` which is described [here](https://www.sharpsightlabs.com/blog/case-when-r/).
> The general syntax is `case_when(condition ~ output-value)`
> 
> For example, a condition would be `treatement == 0` and an output value would be `"Control"`. This would search for every value in the `treatment` column that equals `0` and replace that with the string `"Control"`.


We have started the code for you below. Decide what argument(s) to pass inside the parantheses of `case_when()`.
```{r}
gotv <- gotv |>
  mutate(treatment = case_when()) 
```

Now, when we use `glimpse` we see there is an added `age` variable and that the treatments have word instead of number labels.
```{r}
glimpse(gotv)
```

### Balance table

Next, we're going to confirm that our control and treatment groups look pretty
much the same across a set of covariates, i.e. that the two groups are *balanced on covariates*. Specifically this means we're going to calculate the mean value of a set of covariates across each of the treatment/control
arms, and we expect them to be pretty much equal if our randomization worked. This is related to the idea of exchangeability.

In this exercise, we are going to reproduce a table similar to Table 1 from the paper. We want a table that shows the mean value of the following covariates for each of the five treatment arms: Household size, Nov 2002, Nov 2000, Aug 2004, Aug 2002, Aug 2000, Female, and Age (in years). You should create a table with 5 rows, one for each treatment arm, and 8 columns, one for each covariate of interest. 

We have started some code for you below. What you need to do is:

- Pass an argument to `group_by()` so that we calculate seperate means for each treatment arm.
  - Look [here](https://dplyr.tidyverse.org/reference/group_by.html) for documentation on the `group_by` function.
- Pass an argument to `summarise()` that computes the mean of each covariate in `covariates` for each seperate treatment arm.
  - Look [here](https://dplyr.tidyverse.org/reference/summarise.html) for documentation on the `summarise` function.
  - You may find the function [`across`](https://dplyr.tidyverse.org/reference/across.html) useful here as well. You can use this function *inside* of `summarise()`!

```{r}
covariates <- c("sex", "age", "g2000", "g2002", "p2000", "p2002", "p2004", "hh_size")

gotv_balance <- gotv |>
  group_by(...) |>
  summarise(...)

print(gotv_balance)
```

Note that your numbers will not match up exactly with Table 1. What you want to notice is that the values in each column are similar across the rows.

### Results

Finally, let's replicate the final results (Table 2). For each treatment group, we calculate the percentage of individuals who got out and voted, as well as the total number of individuals in that group! We started the code for you below. You need to 

- Pass an argument to `group_by()` so that we are working with each treatment arm seperately (this should be the same as before!)
- Pass two arguments to `summarise()` to do the following:
  - Create a column titled `Percentage_Voting` that contains the percent of each group that voted
  - Create a column titled `num_of_individuals` that contains the total number of people in that group

```{r}
gotv_results <- gotv |>
  group_by(...) |>
  summarise(...)

print(gotv_results)
```

