---
title: "DiscussionDID"
date: "2023-10-30"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Empirical Application: How the Abolition of Elected Councils Affects Local Public Services in Vietnam

Refer to Tuesday's lecture ([slides](causal3900.github.io/assets/slides/9-1_did.pdf) and recording) for some background on this experiment. Here's a [link](https://www.cambridge.org/core/journals/american-political-science-review/article/impact-of-recentralization-on-public-services-a-differenceindifferences-analysis-of-the-abolition-of-elected-councils-in-vietnam/3477854BAAFE152DC93C594169D64F58) to the paper from the original experiment and a [link](https://www.cambridge.org/core/journals/political-analysis/article/using-multiple-pretreatment-periods-to-improve-differenceindifferences-and-staggered-adoption-designs/747F618FD4AD82A536823521D89310F7) to the paper that re-analyzed this experiment, which we're basing this lab on. 

### Download the data
```{r}
install <- function(package) {
  if (!require(package, quietly = TRUE, character.only = TRUE)) {
    install.packages(package, repos = "http://cran.us.r-project.org", type = "binary")
    library(package, character.only = TRUE)
  }
}

install("cowplot")
install("dplyr")
install("tidyverse")

malesky2014 <- read_csv("https://github.com/causal3900/causal3900.github.io/raw/main/assets/data/malesky2014.csv")
malesky2014 <- drop_na(malesky2014, lnarea, lnpopden, city) # data cleaning
```


### About the data

Variables in the data set:

- `treatment`: binary variable indicating if the record is from the treatment group ($1$) or the control group ($0$)
- `pro4`: educational and cultural program number 4 outcome variable (binary: $1$ if they participated in the program, $0$ if not)
- `tapwater`: tap water outcome variable (binary: $1$ if they drink tap water, $0$ if not)
- `agrext`: agricultural center outcome variable (binary: $1$ if they use the center, $0$ if not)
- `year`: the year the record is from
- `post_treat`: binary variable indicating if the record is from the pre-treatment period ($0$) or the post-treatment period ($0$)

```{r}
head(malesky2014)
```


### Visualizing Trends in the Treatment and Control Groups

In a DID design, the first step is to visualize the trends of the treatment and control groups to check if the parallel trends assumption is credible. Here, we look at three possible outcomes: "Education and Cultural Program," "Tap Water," and "Agricultural Center."

```{r}
# Plot the outcome for `Education and Cultural Program`
ed_and_culture <- malesky2014 %>%
  mutate(treatment = ifelse(treatment == 1, "Treated in 2009","Untreated")) %>%
  group_by(treatment, year) %>%
  summarize(estimate = mean(pro4)) %>%
  ggplot(aes(x = year, y = estimate, color = treatment)) + 
  geom_point() + 
  geom_line() + 
  ggtitle("Ed. and Cultural Program") +
  theme_minimal() +
  theme(legend.position = "none",
        legend.background = element_rect(fill = "white", colour = "white"),
        plot.title = element_text(hjust = 0.5))

# Plot the outcome for `Tap Water`
tap_water <- malesky2014 %>%
  mutate(treatment = ifelse(treatment == 1, "Treated in 2009","Untreated")) %>%
  group_by(treatment, year) %>%
  summarize(estimate = mean(tapwater)) %>%
  ggplot(aes(x = year, y = estimate, color = treatment)) +
  geom_point() +
  geom_line() +
  ggtitle("Tap Water") +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

# Plot the outcome for `Agricultural Center`
ag_center <- malesky2014 %>%
  mutate(treatment = ifelse(treatment == 1, "Treated in 2009","Untreated")) %>%
  group_by(treatment, year) %>%
  summarize(estimate = mean(agrext)) %>%
  ggplot(aes(x = year, y = estimate, color = treatment)) +
  geom_point() +
  geom_line() +
  ggtitle("Agricultural Center") +
  theme_minimal() + 
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
```

Let's compare all three of these plots side-by-side!
```{r}
legend <- get_legend(
  ed_and_culture + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom")
)
pgrid <- cowplot::plot_grid(ed_and_culture, tap_water, ag_center, nrow = 1)
plot_grid(pgrid, legend, ncol = 1, rel_heights = c(1, .1))
```
Notice how we see parallel trends in the pretreatment period for the outcome "Education and Cultural Program," but not for "Tap Water" or "Agricultural Center."

### Basic Difference in Differences Analysis with One Pretreatment Period

```{r}
# remove the extra pretreatment periods from the dataset
onePerMalesky <- malesky2014 %>%
  filter(year==2008 | year ==2010)

# the column treatment I think captures whether you're in treatment vs control group
# the column post_treat I think captures the time--whether you're in pre-treatment or post-treatment period

# did is an interaction term between post_treat and treat
onePerMalesky <- onePerMalesky %>%
  mutate(did = post_treat*treatment)

# regression for Education and Cultural Program on treatment, time period, and interaction term
didReg <- lm(pro4 ~ treatment + post_treat + did, data = onePerMalesky)

summary(didReg)

# estimate
print(didReg$coefficients["did"])
```

### More Advanced DID Design??


