---
title: "Discussion 9 - IV"
date: "2023-10-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Discussion: Instrumental Variables

## Install packages
```{r, echo=FALSE}
install <- function(package) {
  if (!require(package, quietly = TRUE, character.only = TRUE)) {
    install.packages(package, repos = "http://cran.us.r-project.org", type = "binary")
    library(package, character.only = TRUE)
  }
}

install("AER")
install("dplyr")
install("hdm")
```

## Load the data

In this lab we will be examining the following question: does participating
in a 401(k) increase an individual's net wealth? One difficulty in answering
this question is that whether or not people participate in a 401(k) is 
probably not random. However, whether or not somebody is eligible to participate
in a 401(k) is arguably random. Additionally, the only way that 401(k)
eligibility affects an individual's net wealth is by affecting whether or not
they participate in the 401(k). Given this setup, we can use
Instrumental Variables to estimate the Local Average Treatment Effect (average
treatment effect on the compliers) of participating in a 401(k).

Let's begin by loading the data, which will be called `pension`.
```{r}
data(pension)
# Shuffle data around
pension <- pension[sample(1:nrow(pension), nrow(pension)), ]
```

## Set the data up for our analysis

In the data, our variables of interest are the following:
`net_tfa`: This is our outcome variable showing an individual's net wealth.
`p401`: This is our treatment variable showing whether or not an individual participated in a 401(k).
`e401`: This is our instrument showing whether or not an individual was eligible to participate in a 401(k).
`age`, `inc`, `fsize`, `educ`, `db`, `marr`, `twoearn`, `pira`, `hown`: These are
potential confounders, but we will ignore them for right now. Remember we are assuming our instrument is randomly assigned.
```{r}
y <- "net_tfa"
a <- "p401"
z <- "e401"
X <- c("age", "inc", "fsize", "educ", "db", "marr", "twoearn", "pira", "hown")

pension <- select(pension, all_of(...put the variables here...))

glimpse(pension)
```

## How are 401(k) eligibility and participation related

We want to see how 401(k) eligibility and participation are related to build our
intuition for this exercise. In particular it'll be helpful to see how many
people that are eligible to participate in a 401(k) actually do so.
```{r}
table(pension[, c(z, a)])
```

## Estimate the Local Average Treatment Effect

Finally, we will estimate the Local Average Treatment Effect (average
treatment effect on the compliers) of participating in a 401(k) using two-stage
least squares. To do so, we will use the `ivreg` function from the AER package.
To see the help file for this function, run `?ivreg` in your console and take a
look at the documentation.

You will have to create a formula that tells `ivreg` what your outcome variable,
treatment variable, and instrument variable are. **NOTE**: This formula should
look something like `outcome ~ treatment | instrument`.
```{r}
iv_formula <- ...create the forumala here...
iv_model <- ivreg(formula = iv_formula, data = ...)

# Print the model
iv_model
```

The Local Average Treatment Effect is the `p401` coefficient that is in the 
printed summary above!

## Calculate robust standard errors and create 95% conf. interval

Of course, since we want to understand how much uncertainty is in our estimates,
we will calculate the standard errors for our estimate. **NOTE** The `vcov. = vcovHC`
argument below is telling R to calculate robust standard errors, which we typically
want to do in our statistical analyses.
```{r}
standard_err <- summary(..., vcov. = vcovHC)
late <- coef(standard_err)["p401", c("Estimate", "Std. Error")]
# 95% conf. int.
ci <- c(
  late[["Estimate"]] - qnorm(0.975)*late[["Std. Error"]],
  late[["Estimate"]] + qnorm(0.975)*late[["Std. Error"]]
)

# Print the results and confidence interval
cat("LATE (Lower bound, Upper bound):\n",
    paste0("\r", round(late[["Estimate"]], 3)),
    paste0("(", round(ci[[1]], 3), ","),
    paste0(round(ci[[2]], 3), ")\n"))
```
