---
title: "Discussion6-gformula"
author: "Mayleen and Daniel"
date: "2023-09-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
```

# Discussion: Parametric g-formula

## Download the Data

Follow the steps on Ed Discussion to download the data!
```{r}
library(tidyverse)
d <- readRDS("d.RDS")
```

After you download the data and save it into the variable name `d`, go up top to where it says `knitr::opts_chunk$set(echo = TRUE, eval=F)` at the top of the R Markdown file and change `eval` to true, i.e. set `eval=T` so that when you knit the file, your code runs.

## Learn a model to predict $Y$ given $\{A,\vec{L}\}$
You can learn more about the `lm` function [**here**](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/lm).
Note that this function takes several arguments. The first argument is a formula, specifically a symbolic description of the model to be fitted. 

```{r}
# The following formula represents Y as a function of our treatment (A) and all our confounders
f <- y ~ a*(sex + race + mom_educ + dad_educ + log_parent_income +
  log_parent_wealth + test_percentile)

# Estimate the model with OLS
fit <- lm(formula = f, data = d)
```

## Predict conditional average potential outcomes for every unit

Next, we are going to create two copies of our original data. The first copy
will set all the treatment values `= "no_college"` and the second copy will
set all the treatment values `= "college"`.
```{r}
d_all_control <- d %>%
  mutate(a = "no_college")

d_all_treated <- d %>%
  mutate(a = "college")
```

Next we will use our linear regression estimated above to predict both potential
outcomes for each individual using the two data.frames that we just created. We
will then append the estimated potential outcomes to our original data.
To see more details on the `predict` function, see the 
[documentation here](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/predict).

```{r}
# Estimate potential outcomes using our model
potential_outcomes_under_control <- predict(object = fit, newdata = d_all_control)
potential_outcomes_under_treatment <- predict(object = fit, newdata = d_all_treated)

conditional_average_outcomes <- d %>%
  mutate(yhat0 = potential_outcomes_under_control,
         yhat1 = potential_outcomes_under_treatment)
```

## Difference to estimate conditional average effects

Now, we will estimate the causal effect for each individual by taking the
difference between their potential outcomes.

```{r}
conditional_average_effects <- conditional_average_outcomes %>%
  mutate(effect = yhat1 - yhat0)
```

## Average over units

Now, we can calculate the average treatment effect by taking the average
individual effect across the whole data.frame. __HINT__: Use the `summarise`
function to take the mean effect over the whole data.frame.

```{r}
conditional_average_effects %>%
  select(yhat1, yhat0, effect) %>%
  summarise(across(.fns = mean))
```

Alternative:
```
conditional_average_effects %>%
  select(yhat1, yhat0, effect) %>%
  summarise_all(.funs = mean)
```
## Extension 1: Conditional average effects

Modify the procedure above to estimate the average effect in subgroups defined by mom's education:

1.  those with `sex == Male`
2.  those with `sex == Female`

Specifically this means, what is the average effect among those for whom `sex == Male`?
Among those for whom `sex == Female`? __HINT__: this should only be a slight modification
of your code above that takes the average for the whole data.frame.

```{r}
conditional_average_effects %>%
  select(yhat1, yhat0, effect, sex) %>%
  group_by(sex) %>%
  summarise(across(.fns = mean))
```

## Extension 2: Logistic regression

In groups: Repeat the steps above with logistic regression

$$\text{log}\left(\frac{\hat{P}\left(Y\mid A = a, \vec{L} = \vec\ell\right)}{1 - \hat{P}\left(Y\mid A = a, \vec{L} = \vec\ell\right)}\right) = \hat\alpha + A\hat\beta + \vec{L}'\hat{\vec\gamma} + A\vec{L}'\hat{\vec\eta}$$ Helpful hints:

-   This should be a simple extension of all the code above that uses `lm()`.
-   Read about using `glm()` to estimate logistic regression
-   when using `predict()`, search to find out how to predict probabilities

## Extension: Logistic regression

Fit a model

```{r}
f <- y ~ a*(sex + race + mom_educ + dad_educ + log_parent_income +
  log_parent_wealth + test_percentile)

fit <- glm(formula = f, family = binomial, data = d)
```

## Extension: Logistic regression

Predict and summarize to estimate the average effect

\footnotesize

```{r}
potential_outcomes_under_control <- predict(object = fit, newdata = d_all_control, type = "response")
potential_outcomes_under_treatment <- predict(object = fit, newdata = d_all_treated, type = "response")

conditional_average_outcomes <- d %>%
  mutate(yhat1 = potential_outcomes_under_treatment,
         yhat0 = potential_outcomes_under_control,
         effect = yhat1 - yhat0) %>%
  select(yhat1, yhat0, effect) %>%
  summarise(across(.fns = mean))

conditional_average_outcomes
```
