---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Empirical Application: How the Onset of Violent Conflict Affects Economic Output

We will demonstrate the synthetic control method using data from Abadie and 
Gardeazabal (2003), which studied the economic effects of conflict, using the
terrorist conflict in the Basque Country as a case study. This paper used a 
combination of other Spanish regions to construct a synthetic Basque Country
resembling many relevant economic characteristics of the Basque Country before
the onset of political terrorism in the 1970s. The basque data contains
information from 1955–1997 on 17 Spanish regions, including
per-capita GDP (the outcome variable), as well as population density, sectoral
production, investment, and human capital (the predictor variables).

Let's go ahead and load our packages and data. If you notice that there is quite
a lot of missing data, don't worry, that's how it is meant to be!
```{r}
install <- function(package) {
  if (!require(package, quietly = TRUE, character.only = TRUE)) {
    install.packages(package, repos = "http://cran.us.r-project.org", type = "binary")
    library(package, character.only = TRUE)
  }
}

# Load packages
install("dplyr")
install("Synth")

# Load data
data("basque")
glimpse(basque)
```

## Data definitions

The following are definitions of all the variables you see above:

- `regionno` and `regionname`: Numeric identifiers for each Spanish region,
and the names (as a character) of each region.

- `year`: The year corresponding with each row of the data. Spans from
1955-1997.

- `gdpcap`: 1960–1969 averages for real GDP per-capita measured in thousands of
1986 USD.

- Variables with `sec.` prefix: 1961–1969 average percentage of total GDP for
six different industries. For example, the first non-missing value of 
`sec.agriculture` is = 19.54. This means that in 1961, the Agriculture industry 
made up about 20% of the total GDP in all of Spain (note the `regionname`
variable = `Spain (Espana)`.

- Variables with `school.` prefix: 1964–1969 averages for the share of the 
working-age population that was illiterate (`school.illit`), the share with up
to primary school education (`school.prim`), the share with some high school
(`school.med`), the share with high school (`school.high`), and the share with
more than high school (`school.post.high`).

- `popdens`: 1969 population density measured in people per square kilometer.

- `invest`: 1964–1969 averages for gross total investment divided by total GDP.

## Prepare the data for analysis

The first step is to reorganize the dataset into an appropriate format that is
suitable for the main estimator function `synth()`. To do this, we will use the
`dataprep()` function. To see more examples and details on data extraction, run
`?dataprep` in the console below.

In the code below, we need to tell Synth what our predictor variables are, 
as well as how we want to aggregate these variables
(in this case we want to take the average value of them). We also need to tell
synth what time-period we are looking at for these variables (e.g., 1964:1969).
We also need to specify what our outcome variable is (e.g., `gdpcap`) and 
what variable(s) identify the different regions (e.g., `regionname`) and/or 
numbers (e.g., `regionno`).

Finally, we will have to specify what variable is
our time-period variable (e.g., `year`), what region is the treated region
(e.g., region number 17 which is the Basque country), what regions are the
control units (e.g., regions number c(2:16,18)), what is the time-period
over which we want to train our model (e.g., the pre-treatment period
1960:1969), and the time-period over which our outcome data should be plotted
(usually before and after treatment, e.g., from 1955 to 1997).

This is definitely a lot, so please ask us if you have any questions about what
these different variables mean! Additionally, running `?basque` in your console
will give you a pretty nice description of the data.

Okay, now let's prepare what we've just described in code. Below, we've added
comments alongside the lines of code to explain exactly what's happening in
each.
```{r}
dataprep.out <- dataprep(
  foo = basque,          # Our analysis data
  predictors = c(        # This is a list of variables that we want to use as predictors
    "school.illit",      # -   These are the variables that are non-missing for all
    "school.prim",       #     years included in our analysis
    "school.med",
    "school.high",
    "school.post.high",
    "invest"
  ),
  predictors.op = "mean",               # Tells Synth to take the average of all variables in `predictors` above.
  time.predictors.prior = 1964:1969,    # Takes the average of variables in `predictors` from 1964 to 1969.
  special.predictors = list(            # Additional variables to include as predictors in our model:
    list("gdpcap", 1960:1969 , "mean"), # -    Take the average of `gdcap` from 1960 to 1969
                                        # -    Take the average of all others for every OTHER year from 1961 to 1969
    list("sec.agriculture", seq(1961, 1969, 2), "mean"),
    list("sec.energy", seq(1961, 1969, 2), "mean"),
    list("sec.industry", seq(1961, 1969, 2), "mean"),
    list("sec.construction", seq(1961, 1969, 2), "mean"),
    list("sec.services.venta", seq(1961, 1969, 2), "mean"),
    list("sec.services.nonventa", seq(1961, 1969, 2), "mean"),
    list("popdens", 1969, "mean")       # -    Take the average of `popdens` only in 1969
  ),
  dependent = "gdpcap",                 # Specifies our outcome variable
  unit.variable = "regionno",           # Specifies the numeric identifier of each region
  unit.names.variable = "regionname",   # Specifies the name of each region
  time.variable = "year",               # Specifies what variable is our time variable
  treatment.identifier = 17,            # Specifies which region in `regionno` is our treated region
  controls.identifier = c(2:16, 18),    # Specifies which regions in `regionno` should be in our donor pool
  time.optimize.ssr = 1960:1969,        # What years should make up our pre-treatment time period?
  time.plot = 1955:1997                 # What years to plot our outcome variable for?
)
```

Notice that in the code above we use the arguments `predictors`, 
`predictors.op`, and `time.predictors.prior`, and the rest of the information
for the other predictor variables is specified in the `special.predictors`
list. This functionality was designed to allow for easy handling of
several predictors with the same operator (e.g. taking the average) over the
same pre-treatment period (in this case, the school and investment variables)
as well as additional custom (or “special”) predictors with varying operators
and time-periods. For example, the variables for the sector production
shares (e.g. `sec.agriculture`) is only available on a biennial basis
(1961,1963,...,1969) extracted which is why we use the code `seq(1961,1969,2)`.
More details and examples on the use of the `special.predictors`
can be seen by running `?dataprep` in the console.

# Construct our Synthetic Control

Now, we're ready to use the `synth()` function to create the synthetic control
for the GDP of the Basque Country region of Spain. As described in discussion,
this means that `synth()` will create weights for each of the other regions
so that the weighted average of the other regions' GDP will closely match the
true GDP of the Basque Country region.
```{r}
synth.out <- synth(data.prep.obj = dataprep.out, method = "BFGS")
```
For now, we don't need to pay too much attention to the model output. We'll
explore this more below!

## Summarizing our Synthetic Control with Tables

First, we can begin by creating some summary tables from our Synthetic Control
model.
```{r}
synth.tables <- synth.tab(dataprep.res = dataprep.out, synth.res = synth.out)
```

The `synth.tables` variable now contains four tables that will help us evaluate
our synthetic control. The first table looks at our pre-treatment period and
compares the predictor values between the Basque Country (denoted `Treated` in
the table) and our Synthetic Control. **Note** we want the values in the
`Treated` and `Synthetic` columns to be really close together.
```{r}
synth.tables$tab.pred
```
We can see that the values in the `Treated` and `Synthetic` columns are pretty
different for some variables so this would indicate that perhaps our synthetic
control isn't as similar as we would like.

Next, we can look at the weights that got assigned to each of the non-treatment
regions. We can drop regions that have a weight of 0 since those regions don't
contribute to our synthetic control at all!
```{r}
synth.tables$tab.w[synth.tables$tab.w$w.weights != 0, ]
```
From the `w.weights` column we can see that only two regions contribute to our
synthetic control. The Cataluna region makes up approximately 85% of our
synthetic control and the Madrid region makes up the additional 15%. This means
that out of 16 regions in our donor pool, only 2 of them were picked to create
our synthetic control!

Finally we can look at the weights that got assigned to each of our predictor
variables. This can be interpreted as the relative importance of each of our
predictor variables.
```{r}
synth.tables$tab.v
```

We can see that the `school.med`, `school.high`, `school.post.high`, and
`invest` variables had a weight of 0, which means that they are the least
important and had no impact in the creation of our synthetic control. On the 
other hand, the population density in 1969 (`special.popdens.1969`) was the
most important variable.

## Summarizing our Synthetic Control with Plots

Finally, we can plot the economic output for the Basque Contry region and
compare it to the economic output for our Synthetic Control. To make a
convincing case for a large treatment effect, we would like to see the two
trajectories of the outcome variable for the Basque Country and its Synthetic
Control unit to be quite similar prior to the violent conflict and to diverge
sharply when the violent conflict occurs.

Let's create such a plot and see if it indicates a significant treatment effect.
```{r}
path.plot(
  synth.res = synth.out,
  dataprep.res = dataprep.out,
  Ylab = "real per-capita GDP (1986 USD, thousand)",
  Xlab = "year",
  Ylim = c(0, 12),
  Legend = c("Basque country", "synthetic Basque country"),
  Legend.position = "bottomright"
)
abline(a = NULL, b = NULL, h = NULL, v = 1975, col = "red")
```

We can see that the economic output of both units looks super similar until
1975 which is when the violent conflict began in earnest. From that point on,
the economic output of the Basque Country drops significantly. This would
indicate that the violent conflict had a fairly large and negative impact on
economic output in the Basque Country region.

Another way we can visualize this is by creating the same plot, but instead
of showing two lines for the outcome of the Basque Country region and the
outcome of the Synthetic Control Unit, we plot a single line that is the
difference between the two lines in each time period.
```{r}
gaps.plot(
  synth.res = synth.out,
  dataprep.res = dataprep.out,
  Ylab = "gap in real per-capita GDP (1986 USD, thousand)",
  Xlab = "year",
  Ylim = c(-1.5, 1.5),
  Main = NA
)
abline(a = NULL, b = NULL, h = NULL, v = 1975, col = "red")
```

This plot conveys the same information as before. That is, the economic output
of the Basque Country region drops well below the economic output of the
Synthetic Control unit once the violent conflict begins in 1975. In other words
the violent conflict lowers economic output.

## Summary

In this tutorial, we've walked through how to prepare data for the Synthetic
Control method with the following steps:

- Prepare your data using the `dataprep()` function.

- Create a Synthetic Control unit using the `synth` function.

- Evaluate our model with tables using the `synth.tab` function.

- Plot the outcomes of our treated unit and our synthetic control unit with the
`path.plot` and `gaps.plot` functions.

